The development of children's ability to track and predict turn structure in conversation

### 2015-11-28 paper/CasillasFrank-Anticip.tex
completed summary for E2

%We identified anticipatory gaze switches for all 79 usable turn transitions using the same criteria as in Experiment 1. We expected to see the most anticipatory switches in the \textit{normal} condition and the least anticipatory switches in the \textit{no speech} condition because they contained the most and least linguistic information, respectively. We expected to replicate our finding that questions result in more anticipatory switches than non-questions, with the added hypothesis that the question effect is driven by lexicosyntactic cues. We also anticipated an overall increase in anticipatory switches with age. Finally, since the development of prosodic skills partially precedes the development of lexicosyntax, we expected to see more switches in the \textit{prosody only} condition compared to the \textit{words only} condition in the youngest age groups.
%We found that participants at all ages made anticipatory gaze switches more often than would be expected by chance in almost all conditions, but sometimes by a small margin (Figure \ref{fig:randvsrealEN}). For example, anticipatory switches in the \textit{no speech} videos were often modestly above chance (with the exception of 5- and 6-year-olds). One interpretation of this finding is that children do not use linguistic information to predict upcoming turns---the only cue to turn taking in the \textit{no speech} condition was the alternating mouth movements of the conversing puppets. With sufficiently long inter-turn gaps, participants could simply switch their gaze to the responder when the prior talker finishes talking---a strategy that doesn't rely on linguistic information. But this interpretation is not compatible with our the effects of linguistic information that do emerge in the data: An early advantage for prosody and a later advantage for lexical question cues.
%We saw a generalize early advantage for prosody over lexicosyntax in children's spontaneous turn predictions. Notably, while children performed well with \textit{prosody only} speech from the start, they seemed to only reliably exceed chance in the \textit{words only} condition at ages three and older. By age three, the general advantage for prosody over lexicosyntax seems to disappear with more equal rates of switching overall for the two partial linguistic conditions.
%\begin{figure}[t]
%\begin{center}
%\includegraphics[width=0.99\textwidth]{figures/FIG-randvsreal-EN.png}
%\end{center}
%\caption{Baseline-corrected proportion of anticipatory gaze switches made for actual (solid) and randomly-shuffled (dashed) turn transitions in each condition, and across ages. Error bars represent the standard error of the mean.} 
%\label{fig:randvsrealEN}
%\end{figure}
%We observed that, overall, children and adults made the fewest anticipatory switches in the \textit{no speech} videos (9\%) and the most in the \textit{normal} videos (16\%), with the \textit{prosody only} and \textit{words only} videos falling in-between (14\% and 11\%, respectively; Figure \ref{fig:conditionsEN}). However, these condition-based differences were primarily driven by transition type effects (question vs. non-question).
%\begin{figure}[t]
%\begin{center}
%\includegraphics[width=0.65\textwidth]{figures/FIG-conditions-EN.png}
%\end{center}
%\caption{Baseline-corrected proportion of anticipatory gaze switches made for turn transitions in each speech condition, and across ages. Error bars represent the standard error of the mean.} 
%\label{fig:conditionsEN}
%\end{figure}
%As in Experiment 1, we made condition, age and transition type comparisons based on baselin-corrected switching values; we corrected each participant's gaze data by subtracting their random baseline from their actual switching for each turn transition.
%We found that participants made more anticipatory gaze switches following questions than non-questions, but only in the two lexical conditions (\textit{normal} and \textit{words only} speech). The advantage for questions was strongest in the \textit{normal} speech condition, where it emerged early in development. The question effect became larger with age in both lexical conditions. While children made small, sometimes non-linear, gains in their anticipatory switches for non-questions as they grew older, they made large gains in their anticipatory switches with age (especially between ages one and two; Figure \ref{fig:questionsEN}). The increased switching for questions between ages one to two indicates that, around this age, children begin to leverage lexicosyntactic cues in the speech signal to identify questions in ongoing conversation. In the lexical conditions, children had access to early cues to questionhood, such as \textit{wh}-words and subject-auxiliary inversion, to predict an upcoming speaker change. In the \textit{normal} condition, where the question effect was strongest, lexical and syntactic cues were reinforced by question prosody for polar questions.
%\begin{figure}[t]
%\begin{center}
%\includegraphics[width=0.99\textwidth]{figures/FIG-QvsNQ-EN.png}
%\end{center}
%\caption{Baseline-corrected proportion of anticipatory gaze switches made for question- (dashed) and non-question- (dotted) turn transitions in each condition, and across ages. Error bars represent the standard error of the mean.} 
%\label{fig:questionsEN}
%\end{figure}
%Children showed increased anticipatory switching with age across all conditions, making almost as many transitions as adults by ages five and six. Adults reacted to the \textit{no speech} condition differently than the children did, tracking the interaction less closely than in the three linguistic conditions. They may have minimized switching in the \textit{no speech} condition because they realized that there was no conversation to follow.
%\subsubsection{Statistical analysis}
%We tested these findings in two linear mixed-effects models of participants' baseline-corrected anticipatory switches---one model each for children and adults. The child model included condition (\textit{normal}, \textit{words only}, \textit{prosody only}, and \textit{no speech}; \textit{normal} in the intercept), transition type (question vs. non-question), age, and gap duration as fixed effects, with full interactions between condition, transition type, and age. The model also included random effects of item (turn transition) and participant, with random slopes of condition and transition type for participants. The adult model matched the child model, excluding age-related effects.
%There was a significant effect of gap duration. Children had time to make more anticipatory switches in analysis windows with longer gaps (\textit{$\beta$}=0.294, \textit{SE}=0.091, \textit{t}=3.231, \textit{p}$<$.01). An effect of gap duration is expected since prior work using spontaneous anticipatory switching has found that most switches occur in the inter-turn gap (Keitel et al., 2013; Hirvenkari, 2013; Tice and Henetz, 2011).
%The only effects of linguistic condition derived from interactions with age and transition type, despite large differences in the linguistic information available across conditions. (Figure \ref{fig:questionsEN}). The model of children's data confirmed two three-way effects of age, condition, and transition type. Children made significantly more anticipatory switches for questions with age in the \textit{normal} (\textit{$\beta$}=0.04, \textit{SE}=0.008, \textit{t}=5.168, \textit{p}$<$.001) and \textit{words only} conditions (\textit{$\beta$}=0.031, \textit{SE}=0.01, \textit{t}=3.170, \textit{p}$<$.01). Overall, they made a switch twice as often after hearing a question (16\%) than a non-question (8\%) in the \textit{words only} condition, and nearly four times as often in the \textit{normal} condition (19\% vs. 5\%). Since the question effect was limited to the lexical conditions, there was no effect of transition type. There were also no further interactions of age and condition, or of transition type and condition. There was, however, a significant two-way interaction of transition type and age (\textit{$\beta$}=-0.031, \textit{SE}=0.01, \textit{t}=-3.18, \textit{p}$<$.01) in addition to the three-way interactions for the \textit{normal} and \textit{words only} conditions already mentioned.
%Last but not least, we confirmed that children generally made more anticipatory switches as they got older: The model showed a significant effect of age (\textit{$\beta$}=0.028, \textit{SE}=0.009, \textit{t}=3.018, \textit{p}$<$.01). Though the increase in anticipatory looks with age came, in large part, from questions effects in the lexical conditions, children also showed increased switches for non-questions and non-lexical conditions. The increases related to non-questions and non-lexical conditions were overall smaller and sometimes followed a non-linear trajectory (Figure \ref{fig:questionsEN}). As in Experiment 1, early switches (planned before the end of the prior turn) did not show large increases with age, with adults making the fewest early switches in the three linguistic conditions (Figure \ref{fig:cumulativeEN}).
%\begin{figure}[ht]
%\begin{center}
%\includegraphics[width=0.99\textwidth]{figures/FIG-cumulative-EN.png}
%\end{center}
%\caption{Cumulative proportion of anticipatory gaze switches planned during the transition window, across conditions and ages. Age is indicated by hue (darker = older). The solid line is the offset of the pre-gap turn, and the dashed line is the median onset of the post-gap turn in each condition. ``Early switches'' occur before 0.} 
%\label{fig:cumulativeEN}
%\end{figure}
%The advantage for questions over non-questions in these results reinforces our findings from Experiment 1. They suggest that even two-year-olds can reliably distinguish questions from other transition types in third-party conversation, especially when full linguistic information is available. Inspecting the data more closely, we found that 43\% of one-year-olds made more anticipatory switches after questions than non-questions in the \textit{normal} condition, and 24\% of them did so in the \textit{words only} condition. Meanwhile, 61\% of two-year-olds show the effect in the \textit{normal} condition, and 52\% of them do in the \textit{words only} condition, which is already close to the average for older children and adults (Table \ref{tab:questioneffectexp2}).
%\begin{table}[t]
%\begin{center}
%  \begin{tabular}{ccc}
%    \hline
%    Age group & Normal speech & No speech \\ 
%    \hline
%    1 & 0.43 & 0.24 \\
%    2 & 0.61 & 0.52 \\
%    3 & 0.63 & 0.42 \\
%    4 & 0.78 & 0.52 \\
%    5 & 0.91 & 0.52 \\
%    6 & 0.75 & 0.50 \\
%    Adult & 0.67 & 0.57 \\
%    \hline
%  \end{tabular}
%\end{center}
%  \caption{Proportion of participants showing numerically more anticipatory switches for questions than non-questions in the \textit{normal} and \textit{words only} conditions.}
%\label{tab:questioneffectexp2}
%\end{table}
%Finally, we confirmed that young children showed a benefit for prosody over lexicosyntax. We conducted two \textit{post-hoc} \textit{t}-tests for children under three, comparing their anticipatory switching rate against their estimated random switching rate in the \textit{prosody only} and \textit{words only} conditions. We found that anticipatory gazes significantly exceeded chance for the \textit{prosody only} condition (\textit{t}(187.677)=-3.004, \textit{p}=.003), but not for the \textit{words only} condition (\textit{t}(229.828)=-0.8307, \textit{p}=.407). This finding suggests that there is indeed an early advantage for prosody over lexicosyntax in children under three. The advantage for prosody was not driven by a question effect; children and adults did not consistently show more switches after questions than non-questions in the \textit{prosody only} condition (Figure \ref{fig:questionsEN}). So, whereas the early advantage for prosody appeared for both transition types (question and non-question), the later-emerging advantage for lexicosyntax and prosody together was primarily limited to question transitions. This pattern of results may suggest a qualitative shift in the kind of predictions children make while observing a conversation, with linguistic cues (like prosody) playing different roles at different ages.
%The model of adults' anticipatory gaze switches confirmed effects of gap duration and linguistic condition. Longer gaps led to more anticipatory switches (\textit{$\beta$}=0.656, \textit{SE}=0.176, \textit{t}=3.726, \textit{p}$<$.001). Adults also made significantly more anticipatory switches in the \textit{normal} condition than in the \textit{no speech} baseline, indicating that linguistic access did affect their overall performance (\textit{$\beta$}=-0.211, \textit{SE}=0.085, \textit{t}=-2.486, \textit{p}$<$.05). There were no other effects of linguistic condition. The model also showed a marginal two-way interaction of condition and transition type, with adults making more anticipatory gazes for questions in the \textit{normal} condition (\textit{$\beta$}=0.194, \textit{SE}=0.11, \textit{t}=1.773, \textit{p}=.076). Though the question effect was numerically present in both lexical conditions, it was only significant for adults in the \textit{normal} condition, in which they had both lexicosyntactic and prosodic cues to questionhood. This result suggests that prosodic cues still have a role to play in adult question recognition during third-party conversation. Since question effects were primarily limited to the \textit{normal} condition, there was no effect of transition type in the adults' data.

%The core aims of Experiment 2 were to gain better traction on the individual roles of prosody and lexicosyntax in children's turn predictions, and to expand our age range to capture more developmental change. We found that effects of linguistic processing and age were present in this dataset, but both were primarily found within the question effect. Children and adults made more anticipatory switches after questions than non-questions, but did so more often when they had greater access to linguistic information, whether increased access came from age or with speech condition. Our results hint that lexical information is critical in identifying questions during conversation, especially for children. Even so, performance was consistently best in the \textit{normal} condition, in which participants could integrate lexical and prosodic information together.
%By extending the age range of participants, we were able to see more of the developmental trajectory in children's predictions about turn structure. We saw that one-year-olds already made more anticipatory switches than would be expected by chance in two of the four conditions. As they got older, children demonstrated above-chance anticipatory switches in more conditions and made more anticipatory switches overall. Like in Experiment 1, children between ages three and five behaved similarly to each other, and comparably to adults, showing small but steady gains in their anticipatory switches with age. Though the advantage for questions generally increased with age, the biggest developmental shift in our data occurred between one and two years, when the question effect began to reliably emerge in the two lexical conditions. It will be up to future work to disentangle which lexical and prosodic cues children begin to pick out in distinguishing questions from other speech acts, and when.
%The results from Experiment 2 complement those of Experiment 1, but still leave some open questions. For example, we were surprised to find no independent effects of linguistic condition outside of age and transition type interactions. In fact, the primary evidence that participants actually \textit{used} linguistic information came from the question effects, which were limited to the two lexical conditions. Given prior work, it is surprising that taking away lexical and prosodic information didn't have a larger impact on participants' anticipatory looking. We take this issue up further in the general discussion.
%We found that the importance of lexicosyntactic and prosodic information for turn prediction differed depending on age and transition type. While even one-year-olds made anticipatory switches at an above-chance rate in the \textit{normal} and \textit{prosody only} conditions, children didn't consistently surpass chance in the \textit{words only} condition until age three. This pattern of development suggests an early importance for prosodic information in children's turn anticipation. In contrast, the question effect---a primary driver of anticipatory switches in our data---seems to rely on lexical information, with some room for further support from prosody. So, although prosody may help children early on, it does not, on its own, help children to distinguish questions from non-questions. Rather, for both adults and children, prosody's role in the question effect was an additive one: The question effect was strongest when children had access to lexical \textit{and} prosodic information. In short, prosody has different roles to play in turn prediction before and after the lexically-driven question effect emerges. Importantly, these results do not support a straightforward view of lexicosyntax as the primary linguistic cue for turn predictions \citep[even for adults; ][]{de-ruiter2006, magyari2012}, but do support accounts of multiple cue integration \citep{duncan1972, ford1996, hirvenkari2013, torreira2015}.
%We controlled linguistic information in Experiment 2 by using phonetically-manipulated speech, extending speech manipulations used in prior work \citep{de-ruiter2006, keitel2013}. While phonetic manipulation gave us control over the linguistic signal, it also created speech sounds that children don't usually hear in their natural environment. Many prior studies have used phonetically-altered speech with infants and young children \citep[cf.][]{jusczyk2000}, but almost none of them have done so in a conversational context. We found that the \textit{normal} speech condition showed the strongest anticipation effects over all, and we have attributed this to the fact that it gave participants full (lexicosyntactic and prosodic) linguistic information. However, it was also the only natural speech condition. Children could have had trouble processing the \textit{words only} and \textit{prosody only} conditions because they were unfamiliar, and not just because they had less linguistic information available.
%It would take a different type of stimulus to compare lexical and prosodic information without phonetic manipulation. For example, carefully scripted or cross-spliced conversations could control for the presence or absence of prosodic and lexicosyntactic cues to turn transition with natural speech. But since the non-natural (limited cue) linguistic conditions did still elicit some differences in anticipation, children must have processed at least some linguistic information in the \textit{prosody only} and \textit{words only} conditions. Perhaps the advantage of the \textit{normal} condition was then not entirely due to its naturalness. We expect that future work with spontaneous conversational measures will still find an advantage for combined lexical and prosodic cues over lexical cues alone \citep[see also][]{duncan1972, ford1996, torreira2015}.

@@

>>>>

Children and adults both showed more anticipatory gaze switches with increased linguistic information, but only for a subset of the linguistic conditions and transition types. We had expected to see the most anticipatory switches in the \textit{normal} condition and the least anticipatory switches in the \textit{no speech} condition because they contained the most and least linguistic information, respectively. We had also expected to replicate our finding from Experiment 1 that questions result in more anticipatory switches than non-questions, with the added hypothesis that the question effect is driven by lexicosyntactic cues. We additionally anticipated an overall increase in anticipatory switches with age. Finally, since the development of prosodic skills partially precedes the development of lexicosyntax, we expected to see more switches in the \textit{prosody only} condition compared to the \textit{words only} condition in the youngest age groups.

In fact, children and adults did show more anticipatory gaze switching in the normal condition compared to the no-speech condition, but for children this effect only emerged in the form of anticipations following question turns, which increased with age faster than it did with no linguistic cues at all (i.e., in the no-speech condition). Adults also showed significantly higher anticipatory switch rates in the words only condition but no effects of transition type, alone or within linguistic conditions. Taken together, these results partly replicate the findings from Experiment 1: participants make more anticipatory switches when they have access to lexical information and, when they do, tend to make more anticipatory switches for questions compared to non-questions.

We had anticipated significant gains in anticipatory switching with age, but children only showed significant developmental increases in the prosody only condition and the normal condition (for question transitions). Rather than showing an early advantage for prosody over lexical information, children did not show significant improvement (or even above-chance performance) until age five. In contrast, their anticipatory gaze switches were, on average, already above chance at age one in the normal condition (with both lexical and prosodic information) and at age three in the words only condition (with lexical information). These findings do not support an early role for prosody in children's spontaneous predictions about upcoming turn structure. On the contrary, their predictions were best when lexical information was present, especially when following a question.

However, these results do not support the idea that lexical information is sufficient for children (as is proposed for adults; \citealp{de-ruiter2006}). On the contrary, children showed significant gains in the normal speech condition, but \textit{not} in the words only condition. Notably, the normal speech condition, in addition to having both lexical and prosodic information, is also the one most likely to occur in our participants' daily lives (compared to muffled or robotic speech) and therefore may have an extra advantage over the other conditions.

Finally, on average, participants at all ages made anticipatory gaze switches more often than would be expected by chance in the \textit{no speech} condition. One interpretation of this finding is that participants were not using linguistic information to predict upcoming turns at all---the only cue to turn taking in the \textit{no speech} condition was the alternating mouth movements of the conversing puppets. But this interpretation is not compatible with the effects of linguistic information that do emerge in the adult and child data: An advantage for lexical cues shaped by transition type and age.

The core aims of Experiment 2 were to gain better traction on the individual roles of prosody and lexicosyntax in children's turn predictions, and to expand our age range to capture more developmental change. We found that effects of linguistic processing and age \textit{were} present in the dataset, but were primarily related to the question effect, and primarily occur in the normal speech condition. The relation between prosody and lexicosyntax in young children's anticipations is still not clear: The results of the prosody only manipulation suggest that children do not reliably use low-pass filtered speech to anticipate upcoming turn structure until age five, but on the other hand, the words only condition did not show significant overall differences from the no-speech baseline, including its interactions with age or transition type. If anything, the results do not support the idea that there is a strong early advantage for prosody in turn predictions.

@@

%Though children's anticipatory gazes did show some gains with age, adults on average still made more anticipatory switches than children in all conditions, even compared to 6-year-olds. Curiously, child and adult performance were much more similar by ages 3--5 in Experiment 1. This may indicate that children rely more in non-verbal cues in anticipating turn transitions or, alternatively, that adults are better at flexibly adapting to the turn-relevant cues present at any moment.

%We controlled linguistic information in Experiment 2 by using phonetically-manipulated speech, extending speech manipulations used in prior work \citep{de-ruiter2006, keitel2013}. While phonetic manipulation gave us control over the linguistic signal, it also created speech sounds that children don't usually hear in their natural environment. Many prior studies have used phonetically-altered speech with infants and young children \citep[cf.][]{jusczyk2000}, but almost none of them have done so in a conversational context. We found that the \textit{normal} speech condition showed the strongest anticipation effects over all, and we have attributed this to the fact that it gave participants full (lexicosyntactic and prosodic) linguistic information. However, it was also the only natural speech condition. Children could have had trouble processing the \textit{words only} and \textit{prosody only} conditions because they were unfamiliar, and not just because they had less linguistic information available.
%
%It would take a different type of stimulus to compare lexical and prosodic information without phonetic manipulation. For example, carefully scripted or cross-spliced conversations could control for the presence or absence of prosodic and lexicosyntactic cues to turn transition with natural speech. But since the non-natural (limited cue) linguistic conditions did still elicit some differences in anticipation, children must have processed at least some linguistic information in the \textit{prosody only} and \textit{words only} conditions. Perhaps the advantage of the \textit{normal} condition was then not entirely due to its naturalness. We expect that future work with spontaneous conversational measures will still find an advantage for combined lexical and prosodic cues over lexical cues alone \citep[see also][]{duncan1972, ford1996, torreira2015}.

%%%%%%%
